Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-28 main_new.tex
Trygve updates

\begin{itemize}
\item Full numerical ocean models cannot be run online on the sensing
  platforms due to their computational load for accurate results,
  hence statistical proxy models of the environment must be used.
\item Decisions on where to sample happen sequentially, i.e.  recently
  acquired information can be assimilated into and used to improve
  statistical models and reduce uncertainty associated with where to
  make measurements.
\item As a-priori information about the state of the ocean is very
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}; this is
  usually referred to as the \emph{adaptivity gap}
  \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the most resource-efficient design to collect necessary
  data), and computability (arriving at a solution in reasonable
  time). To successfully resolve features, this trade-off has to be
  considered in sample design and practice.
\end{itemize}

@@

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network, where robotic
  assets augment ship-based measurements to increase coverage and
  resolution of phenomena of scientific interest.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more autonomous and collaborative networked operations in
    order to resolve the numerous bio-geochemical processes and their
    interactions.}
\label{fig:envir}
\end{figure}

@@

\begin{figure}[!h] 
\centering 
\includegraphics[width=0.85\textwidth]{Figures/pictures/c-updated.pdf}
\caption{Frontal patterns off of the Nidelva river, Trondheim, Norway.}
\label{fig:nidelven}
\end{figure}

@@

The simplest heuristic for adaptive sampling is to choose the next
sampling location based on current EPs. The largest variance in
sampling locations is an indication of higher uncertainty; therefore
one selects the design location with an EP closest to $0.5$.

At stage $s$, based on the currently available data
$\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from
Eq. (\ref{gp_upd}), with mean $\bm_{s-1}=E(\bxi|\mathcal{Y}^{s-1})$
and covariance matrix $\bA_{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$.
The next round of measurements $\by^{j,s}$, can be gathered with
design $d^{j,s}$, $j=1,\ldots,J$, where $J$ indicates all possible AUV
trajectories from the current stage. The {\it{naive}} strategy then,
selects the design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature
or salinity variables, only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Neither does it
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced and therefore
susceptible to local minimas.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more round of measurements. In this selection
strategy there is no anticipation of what the subsequent designs might
offer, beyond the first round.

@@

Note that this strategy gives a sequential conditional version of the
formula in Eq. (\ref{sur}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is with respect to the conditional density
$\pi(\by^{j,s}|\mathcal{Y}^{s-1})$. A similar closed form calculation
for expected IBV is hence applicable in Eq. (\ref{critSEQ}), using the
updated GP model from step $s-1$.  Once the data is collected for the
best design, the GP model is updated again. The mean $\bm_{s}$ and
covariance matrix $\bA_{s}$ are used to compute the next design at
stage $s+1$, and so on.

Even though this myopic strategy is non-anticipative, it still gives a
reasonable approach for creating designs in many
applications. Moreover, it is easily implemented on-board an AUV,
using the efficient approach for data updating of the GP model and the
calculation of the closed form expected IBV expressions for each
subsequent trajectory.

@@

We now extend the myopic strategy to a look-ahead strategy which is
optimal when one can gather only two more rounds of measurements. In
addition to the next round of measurements $\by^{j,s}$, this
look-ahead strategy, anticipates the subsequent design $j_2$ with
data $\by^{j_2,s+1}$, when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ U_{la,\mbox{upd,j}} \right\},  \\
    U_{la,\mbox{upd},j} & = &  E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ \mbox{argmin}_{j_2 \in \{1,\ldots,{J}_2\}} \left[ V_{la,\mbox{upd},j_2} \right] \right\}, \nonumber \\
V_{la,\mbox{upd},j_2} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j_2,s+1}|\mathcal{Y}^{j,s}} \left\{ p(\bx;\mathcal{Y}^{j,s+1})\left( 1-p(\bx;\mathcal{Y}^{j,s+1})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s+1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by^{j_2,s+1},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by^{j,s},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by^{j_2,s+1},\mathcal{Y}^{j,s}\}$ represent
the sets of data variables, and the expectations are with respect to
the conditional densities $\pi(\by^{j,s}|\mathcal{Y}^{s-1})$ and
$\pi(\by^{j_2,s+1}|\by^{j,s},\mathcal{Y}^{s-1})$.

Because of the intermixed optimization and expectation, there is no
longer a closed form for the required integrals. Instead, we solve the
first expectation by Monte Carlo sampling of data $\by^{j,s}$ from its
conditional distribution. For each data sample, the second expectation
is solved using the closed form expressions for expected IBV from
Section \ref{sec:sur}.

Even though the strategy looks at two stages, it is only used to find
the current best design. When data is collected, the GP model is
updated, and the mean $\bm_{s}$ and covariance matrix $\bA_{s}$ are
used to compute the next design at stage $s+1$, now anticipating what
stage $s+2$ could offer, and so on.

This look-ahead approach is much more computationally demanding than
the myopic strategy, and for practical implementation we prune paths
in the evaluation of Eq. (\ref{critLA}). This means that we do not
compute all possible branches of the first two stages, as they are
indicated in Figure \ref{fig:PathSelOpt}. Instead, we use the myopic
strategy to rank the three best designs on the first stage alone, and
for each of these we undertake the look-ahead calculations.

>>>>

\begin{itemize}
\item Full numerical ocean models cannot provide accurate results online if run on robotic sensing
  platforms, as on board computers cannot deliver the computational power required,
  hence statistical proxy models of the environment must be used instead.
\item Decisions on where to sample happen sequentially, i.e.  recently
  acquired information can be assimilated into and used to improve
  statistical models and reduce uncertainty associated with where to
  make measurements.
\item As a-priori information about the state of the ocean is very
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}; this is
  usually referred to as the \emph{adaptivity gap}
  \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization
  (finding the most resource-efficient design to collect necessary
  data), and computability (arriving at a solution in reasonable
  time). To successfully resolve features, this trade-off has to be
  considered in sample design and practice.
\end{itemize}

@@

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network, where robotic
  assets augment ship-based measurements to increase coverage and
  resolution of phenomena of scientific interest.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more autonomous and collaborative networked operations in
    order to resolve the numerous bio-geochemical processes and their
    interactions.}
\label{fig:envir}
\end{figure}

@@

%\begin{figure}[!h] 
%\centering 
%\includegraphics[width=0.85\textwidth]{Figures/pictures/c-updated.pdf}
%\caption{Frontal patterns off of the Nidelva river, Trondheim, Norway.}
%\label{fig:nidelven}
%\end{figure}

@@

The simplest heuristic for adaptive sampling is to choose the next
sampling location based on current EPs. The largest variance in
sampling locations indicates higher levels of uncertainty; therefore
one selects the design location with an EP closest to $0.5$.

At stage $s$, based on the currently available data
$\mathcal{Y}^{s-1}$, we fit an updated Gaussian model from
Eq. (\ref{gp_upd}) with mean $\bm_{s-1}=E(\bxi|\mathcal{Y}^{s-1})$
and covariance matrix $\bA_{s-1}=\mbox{Var}(\bxi|\mathcal{Y}^{s-1})$.
The next round of measurements $\by^{j,s}$ can be gathered with
design $d^{j,s}$, $j=1,\ldots,J$, where $J$ indicates all possible AUV
trajectories from the current stage. The {\it{naive}} strategy, then,
selects the design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} |p(\bx_{d^{j,s}};\mathcal{Y}^{s-1})-0.5|, \\
    p(\bx;\mathcal{Y}^{s-1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S | \mathcal{Y}^{s-1}). \nonumber
\end{eqnarray}
This strategy does not account for the uncertainty in the temperature
or salinity variables, but is instead applicable only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Additionally, this strategy does not
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced, and is therefore
susceptible to local minimas.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more round of measurements. In this selection
strategy, there is no anticipation of what the subsequent designs might
offer beyond the first round.

@@

Note that this strategy gives a sequential conditional version of the
formula in Eq. (\ref{sur}). Now $\mathcal{Y}^{s-1}$ is available, and
the expectation is with respect to the conditional density
$\pi(\by^{j,s}|\mathcal{Y}^{s-1})$. A similar closed-form calculation
for expected IBV is hence applicable in Eq. (\ref{critSEQ}) using the
updated GP model from step $s-1$.  Once the data is collected for the
best design, the GP model is updated again. The mean $\bm_{s}$ and
covariance matrix $\bA_{s}$ are used to compute the next design at
stage $s+1$, and so on.

Even though this myopic strategy is non-anticipative, it still gives a
reasonable approach for creating designs in many
applications. Moreover, it is easily implemented on-board an AUV,
using an efficient approach for data updating the GP model and the
calculation of the closed-form expected IBV expressions for each
subsequent trajectory.

@@

We now extend the myopic strategy to a look-ahead strategy, which is
optimal when one can gather only two more rounds of measurements. In
addition to the next round of measurements $\by^{j,s}$, this
look-ahead strategy, anticipates the subsequent design $j_2$ with
data $\by^{j_2,s+1}$ when choosing the current design $d^{j,s}$.  The
selected design is:
\begin{eqnarray}\label{critLA}
    d^{*,s} &=& \mbox{argmin}_{j \in \{1,\ldots,J\}} \left\{ U_{la,\mbox{upd,j}} \right\},  \\
    U_{la,\mbox{upd},j} & = &  E_{\by^{j,s}|\mathcal{Y}^{s-1}} \left\{ \mbox{argmin}_{j_2 \in \{1,\ldots,{J}_2\}} \left[ V_{la,\mbox{upd},j_2} \right] \right\}, \nonumber \\
V_{la,\mbox{upd},j_2} & \approx & \sum_{\bx \in \mathcal{M}_g} E_{\by^{j_2,s+1}|\mathcal{Y}^{j,s}} \left\{ p(\bx;\mathcal{Y}^{j,s+1})\left( 1-p(\bx;\mathcal{Y}^{j,s+1})\right) \right\} \Delta, \nonumber \\
    p(\bx;\mathcal{Y}^{j,s+1}) &=& P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by^{j_2,s+1},\mathcal{Y}^{j,s}). \nonumber
\end{eqnarray}
Here, $\mathcal{Y}^{j,s}=\{\by^{j,s},\mathcal{Y}^{s-1}\}$ and
$\mathcal{Y}^{j,s+1}=\{\by^{j_2,s+1},\mathcal{Y}^{j,s}\}$ represent
the sets of data variables, and the expectations are calculated with respect to
the conditional densities $\pi(\by^{j,s}|\mathcal{Y}^{s-1})$ and
$\pi(\by^{j_2,s+1}|\by^{j,s},\mathcal{Y}^{s-1})$.

Because of the intermixed optimization and expectation, there is no
longer a closed form for the required integrals. Instead, we solve the
first expectation by Monte Carlo sampling of data $\by^{j,s}$ from its
conditional distribution. For each data sample, the second expectation
is solved using the closed-form expressions for expected IBV from
Section \ref{sec:sur}.

Even though the strategy looks at two sampling stages, it is only used to find
the current best design. When data are collected, the GP model is
updated, and the mean $\bm_{s}$ and covariance matrix $\bA_{s}$ are
used to compute the next design at stage $s+1$, now anticipating what
stage $s+2$ could offer, and so on.

This look-ahead approach is much more computationally demanding than
the myopic strategy, so for practical implementation, we prune paths
in the evaluation of Eq. (\ref{critLA}). This means that we do not
compute all possible branches of the first two stages, as they are
indicated in Figure \ref{fig:PathSelOpt}. Instead, we use the myopic
strategy to rank the three best designs on the first stage alone, and
for each of these we undertake the look-ahead calculations.
